---
title: "Case study: mtDNA haplogroups and HIV status in Botswana"
format:
  html:
    toc: true
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(here)
```

> Work a real dataset end-to-end: from a sequencing-derived haplogroup call to a
> hypothesis test — and to the reason the hypothesis test cannot be trusted.

## Why this module

Every dataset you have analysed so far in this series was chosen because it
behaves. This one does not. It is a real study: 338 participants from Botswana,
mitochondrial genomes sequenced, haplogroups assigned with
[HaploGrep 3](https://haplogrep.i-med.ac.at/), and paired with HIV status.

The scientific question was simple: **does mitochondrial ancestry associate with
HIV status?**

The answer turns out to be *"we cannot tell from this data"* — and finding out
*why* is the actual lesson. Learning to recognise that situation is worth more
than another clean p-value.

## Learning objectives

By the end of this module you should be able to:

1. Load and inspect a real haplogroup dataset with the `tidyverse`.
2. Derive a coarser grouping (macro-haplogroup) from a fine-grained label.
3. Build and interpret a contingency table.
4. Choose between a chi-squared and a Fisher exact test, and know why.
5. **Detect confounding before you believe a result.**

## The data

::: {.callout-important}
## This is a synthetic teaching dataset
The real Botswana cohort is **not** distributed with this workshop — it carries
participant HIV status and stays private. The block below regenerates a
**synthetic** dataset with the same shape (338 participants, L-lineage
haplogroups) and, crucially, the same fatal design flaw: sex perfectly
confounded with HIV status. Every number on this page comes from that synthetic
data. Do not cite any of it as a real epidemiological result.
:::

```{r}
#| message: false
set.seed(1)
n <- 338

# Macro-haplogroups: L0 dominant, L1 rare — as in the real Southern African cohort
macro <- sample(c("L0", "L1", "L2", "L3"), n, replace = TRUE,
                prob = c(0.53, 0.05, 0.16, 0.24))

# Fine-grained labels, long-tailed so most have only a handful of carriers
sub_pool <- c("a1","a1b","a1b1","a2","a2a","a2a1","a2a2","a2a2a","b1","b2",
              "b2a","b2b","c","c1","d","d1","d1a","d1b","d1b1","d1b2",
              "d1b2b","d1b2b1","d1b2b1b","d3","d3a","d3b","e","e1","e1a",
              "e1a1","e1a1a","f","f1","f1a","f1b","f1b1","g1","h","h1","k1a")
haplogroup <- paste0(macro,
  sample(sub_pool, n, replace = TRUE, prob = rev(seq_along(sub_pool))^1.3))

# The sampling artifact: every man entered through the HIV-negative arm, so the
# HIV-positive group is women only. HIV status is otherwise independent of
# haplogroup — there is no real haplogroup effect built in.
sex <- sample(c("F", "M"), n, replace = TRUE, prob = c(0.88, 0.12))
hiv_status <- ifelse(sex == "M", "negative",
                     sample(c("positive", "negative"), n,
                            replace = TRUE, prob = c(0.75, 0.25)))

mtdna <- tibble(
  participant_id   = sprintf("BW_%04d", sample(n)),
  haplogroup       = haplogroup,
  macro_haplogroup = macro,
  sex              = sex,
  hiv_status       = hiv_status
)
glimpse(mtdna)
```

Five columns:

| Column | Meaning |
|---|---|
| `participant_id` | De-identified participant code |
| `haplogroup` | Full HaploGrep 3 call, e.g. `L0d1b2b1b` |
| `macro_haplogroup` | First two characters, e.g. `L0` |
| `sex` | Recorded sex |
| `hiv_status` | `positive` / `negative` |

::: {.callout-note}
## Where `macro_haplogroup` comes from
Haplogroup labels are hierarchical: `L0d1b2b1b` sits inside `L0d1b2`, inside
`L0d`, inside `L0`. Truncating to two characters collapses the
`r n_distinct(mtdna$haplogroup)` fine-grained labels here into
`r n_distinct(mtdna$macro_haplogroup)` macro-groups. We do this because most of
those labels have only a handful of participants — far too few to test.
:::

## Step 1 — Look before you test

```{r}
mtdna |> count(macro_haplogroup, sort = TRUE)
```

Four macro-haplogroups, all within the L lineage — as expected for a Southern
African cohort. `L0` dominates, which is consistent with L0 being the deepest
branch of the human mtDNA tree and common in Khoe-San-related ancestry.

```{r}
#| fig-width: 7
#| fig-height: 4
mtdna |>
  count(macro_haplogroup) |>
  ggplot(aes(x = fct_reorder(macro_haplogroup, -n), y = n)) +
  geom_col(fill = "#4C72B0") +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  labs(x = "Macro-haplogroup", y = "Participants",
       title = "mtDNA macro-haplogroup distribution (n = 338)") +
  theme_minimal(base_size = 12)
```

## Step 2 — The contingency table

```{r}
tab <- table(mtdna$macro_haplogroup, mtdna$hiv_status)
tab
```

Eyeball it before testing. Roughly two-thirds of every row is `positive`. The
rows look *similar* — which is already a hint about what the test will say.

```{r}
#| fig-width: 7
#| fig-height: 4
mtdna |>
  ggplot(aes(x = macro_haplogroup, fill = hiv_status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c(negative = "#8FBCD4", positive = "#C44E52")) +
  labs(x = "Macro-haplogroup", y = "Proportion", fill = NULL,
       title = "HIV status within each macro-haplogroup") +
  theme_minimal(base_size = 12)
```

## Step 3 — Test it

```{r}
chisq.test(tab)
```

The p-value is well above 0.05 — no evidence of association.

::: {.callout-tip}
## chi-squared or Fisher?
Chi-squared is an approximation that degrades when expected cell counts fall
below ~5. Check first:

```{r}
chisq.test(tab)$expected |> round(1)
```

The `L1` row is thin. Fisher's exact test does not rely on the approximation:

```{r}
fisher.test(tab, simulate.p.value = TRUE, B = 10000)
```

Both agree — no association. When two tests with different assumptions agree,
you can stop worrying about which one to report.
:::

## Step 4 — The part that matters {#sec-confounding}

Before you write "no association between mtDNA haplogroup and HIV status" in a
manuscript, run one more line:

```{r}
table(mtdna$sex, mtdna$hiv_status)
```

Read that table carefully.

**Every single male in this cohort is HIV-negative. Every HIV-positive
participant is female.**

That is not biology. No cohort in Botswana looks like this by chance — HIV
prevalence in men is not zero. This is a **sampling artifact**: the
HIV-negative comparison group was recruited in a way that pulled in the only
men in the study.

The consequence is severe. Sex is perfectly confounded with HIV status, so:

- Any haplogroup–HIV signal is inseparable from a haplogroup–sex signal.
- You cannot adjust for sex. Adjustment needs *some* overlap — HIV-positive men
  or HIV-negative-only women to compare against. There are none.
- The null result above is therefore **uninformative**, not reassuring. We did
  not find "no effect"; we found that this design cannot measure the effect.

::: {.callout-important}
## The transferable lesson
A p-value answers the question you asked. It cannot tell you the question was
unanswerable. **Always cross-tabulate your exposure against your key covariates
before you interpret a test.** One `table()` call would have saved this analysis.

Because mtDNA is maternally inherited, a sex imbalance is *especially*
dangerous here — but this failure mode is completely general.
:::

## Exercises

**1.** Restrict the analysis to women only and re-run the test. What is the
p-value now? Does restricting fix the confounding, and what did you pay for it?

```{r}
#| eval: false
women <- mtdna |> filter(sex == "F")
# your code here
```

**2.** Collapse to `L0` vs everything else and test with `fisher.test()`. Does a
binary comparison change the conclusion?

**3.** The fine-grained haplogroups were collapsed to 4 macro-groups. Count
participants per full `haplogroup`. How many have fewer than 5? What would
happen to a chi-squared test run on the uncollapsed table?

**4.** *Discussion.* You are handed this dataset and asked to publish the null
result. What do you tell the team, and what would you need to collect to
answer the original question properly?

## What to take away

- Real data arrives with defects that no amount of correct statistics repairs.
- Look at marginals and cross-tabs **before** testing.
- A null result from a confounded design is not a null result.
- Reporting *why* a question cannot be answered is a legitimate scientific
  contribution.

::: {.callout-note collapse="true"}
## About this dataset
**Synthetic.** It is generated in the code block above to mirror a real study of
mitochondrial genomes in Botswana — the same size, the same L-lineage
haplogroup structure, and the same sex/HIV confounding that makes the real
analysis uninterpretable — without using any real participant data. The real
cohort is not distributed: it carries HIV status and remains private. Nothing on
this page should be read as an actual result from the Botswana study.
:::
